---
title: "R Fridays GIS Tutorial"
author: "Danielle Clake"
date: "12 March 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = TRUE)
```

```{r, eval=FALSE}
setwd("C:/User/Documents/R Fridays/GIS Data in R")
```

#Spatial Data Types
The two major types of GIS data are *vector* format (a set of spatially referenced points, lines, or polygons that are each assigned one or more attributes) and *raster* format (a continuous grid of pixels or "cells" which each have an assigned value - think of a digital image as an example, where the assigned value would be a specific colour for each pixel). We will use both in this tutorial.

#Install and Load Packages
This code will install and load the packages used in this tutorial. If packages are already installed and/or loaded users should skip the code below. Prompts to load packages will also be provided in text before the first time each package is used.
```{r}
x <- c("raster", "rgdal", "rgeos", "tmap")
```

Uncomment the code below if you need to install/load packages:
```{r, results = 'hide', warning = FALSE, message = FALSE}
#install.packages(x)
lapply(x, library, character.only = TRUE)
```


#Vector Data
##Load data and view structure
We will be using the **rgal** package to load our spatial data in R. This package has a function *readOGR* which requires two inputs: dsn (the "data source name") and the name of the layer. Note that for this function we do not need to include a file extension after the data name.
```{r, results = 'hide', warning = FALSE, message = FALSE}
library(rgdal)
PROVINCE <- readOGR(dsn = "Data/shapefiles", layer = "Prov_Boundary_GeoGratis")
plot(PROVINCE) #(sorry territories - lots of small islands made it slow to plot)
```

The information for this spatial data is stored in "slots":
```{r}
#str(PROVINCE)
slotNames(PROVINCE)
```

We can use the "@data" slot to see the attribute table for our data. Using the "$" will automatically query for that column name in the data.
```{r}
PROVINCE@data
PROVINCE$NAME

#To create a new shapefile that only includes Alberta:
AB <- PROVINCE[PROVINCE$NAME == "ALBERTA",]
plot(AB)
```

The "@proj4string" slot tells us the coordinate reference system (CRS) for this data.
```{r}
AB@proj4string
```

I won't go into too much detail about CRS and projections today, but we do need to make sure that we are using the same projection. The province layer is in a geographic coordinate system using lat/long and decimal degrees as units.
```{r, results = 'hide', warning = FALSE, message = FALSE}
#If we add data that is in a different CRS (e.g. UTM Zone 12N), it will not line up properly
RSA_U12 <- readOGR(dsn = "Data/shapefiles", layer = "RSA_UTM12")
RSA_U12@proj4string
plot(RSA_U12, add = TRUE)

#We can reproject it to match our province CRS as follows:
RSA <- spTransform(RSA_U12, proj4string(AB))
plot(RSA, add = TRUE)
```


##Add sampling points
Often times, we will have a table with coordinates that we need to turn into spatial data
```{r}
#Import coordinate table
SL <- read.csv("./Data/SampleLocations_UTM12.csv")

#Turn into a "Spatial Points Data Frame" by specifying columns with coordinates
coordinates(SL) <- cbind("UTM_East", "UTM_North")

#Specify projection
proj4string(SL) <- "+proj=utm +zone=12 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

str(SL)
plot(SL)
```


##Add and view sample data
Now that we have our pretend locations, we can pretend that we collected some data at these locations, and add it to our points.
```{r}
set.seed(12)
SL$Abundance <- rpois(12, 5)
SL@data
```

Next, use the *tmap* package to visualize these data
```{r, results = 'hide', warning = FALSE, message = FALSE}
#First, add our local study area (LSA) to add context to our points
LSA_U12 <- readOGR(dsn = "Data/shapefiles", layer = "LSA_UTM12")

#Now create a quick map
Abun_Map <- tm_shape(LSA_U12) +
  tm_polygons(col = "white") +
tm_shape(SL) +
  tm_bubbles("Abundance", 
             col = "blue", 
             scale = 2, 
             sizes.legend = c(1, 5, 10)) +
  tm_text("Loc_ID", auto.placement = 1) +
tm_layout(outer.margins=0, inner.margins=c(0.15,0.15,0.05,0.05), asp = 0) +
tm_scale_bar()

Abun_Map
```

##Basic spatial analysis
We can then do various analyses on these spatial points, for example by calculating the distance between them using the "spDists" tool in the *sp* package:
```{r}
spDists(SL)
```

Perhaps we also want to know how long it would take to visit each of these locations. We can do this by creating a line between them, then calculating the length of that line.
```{r}
#Create line that passes through all sample locations
PathLine <- SpatialLines(list(Lines(Line(cbind(SL$UTM_East,SL$UTM_North)), "L1")))

#Define projection for the line
proj4string(PathLine) <- "+proj=utm +zone=12 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Plot line on top of sample locations
Abun_Map +
  tm_shape(PathLine) +
    tm_lines(col = "black")

#Calculate length of line (the units here will be meters based on our projection)
SpatialLinesLengths(PathLine)
```


##Add landcover data as vector
In the "data" folder I have included a clipped land cover dataset downloaded from the Alberta Biodiversity Monitoring Institute (ABMI)
```{r, results = 'hide', warning = FALSE, message = FALSE}
ABMI_LC <- readOGR(dsn = "Data/shapefiles", layer = "LC_ABMI_2010_UTM12")
head(ABMI_LC@data)
```

We can also map this land cover data. Here I have picked certain colours in a palette to represent the different land cover types. Since there were relatively few classes I picked them manually using the following website:
https://htmlcolorcodes.com/

However there are also some good "cheatsheets" and programs available to do this automatically:
https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf

```{r}
#Create colour palette for data
levels(ABMI_LC$LC_desc)
LC.palette <- c("#3e7a3e", "#e2825f", "#ad8048", "#cff94f", 
                "#a9adaa", "#f7cb5b", "#ffffff", "#6cbdf7")

#Create map of data:
tm_shape(ABMI_LC) + #Specify shapefile to be mapped
  tm_fill("LC_desc", palette = LC.palette,  title = "Land Cover") + #select fill data
tm_shape(SL) +
  tm_bubbles("Abundance", 
             col = "black", 
             scale = 1, 
             sizes.legend = c(1, 5, 10)) +
tm_layout(outer.margins=0, inner.margins=c(0.15,0.3,0.01,0.01), asp = 0) +
tm_scale_bar()
```


#Raster Data
Often the spatial data that we want to use is best represented in raster format - especially for continuous variables (e.g. elevation) or variables covering a landscape (e.g. landcover)

First, create a template raster with the spatial extent ("ext") and projection ("crs") set to be the same as our polygon landcover layer ("ABMI_LC"). We are using a resolution, or pixel size, of 100 map units (in this case metres).
```{r}
template.raster <- raster(ext = extent(ABMI_LC), resolution = 100, crs = CRS(projection(ABMI_LC)))
```

Next we essentially "stamp" our land cover polygons onto our template raster.
```{r}
#This step may take a couple minutes to run:
LCrast <- rasterize(x = ABMI_LC, y = template.raster, field = "LC_class", fun = 'last') 

plot(LCrast) #lets see what our raster looks like!
```

##Buffer sampling locations and extract values
First we need to create "buffers", or circles with a set radius around our sampling locations. We will do this using the *gBuffer* tool in the **rgeos** package. We then can extract the landcover values within each buffer using the *extract* tool in the **raster** package.
```{r}
#Create 1000 m buffers around sampling locations:
SL_buffer1km <- gBuffer(spgeom = SL, byid = TRUE, id = SL$Loc_ID, width = 1000)
plot(LCrast)
plot(SL_buffer1km, add = TRUE) #look at the buffers created

#Extract number of pixels of each land cover class under each polygon:
LC1km <- raster::extract(x = LCrast, y = SL_buffer1km)

#Find frequency of each land class within each polygon:
LC1km.fq <- lapply(LC1km, table) 

#Calculate proportion of land cover within each polygon:
LC1km.pr <- lapply(LC1km.fq, FUN = function(x){x/sum(x)}) 

LC1km.pr
```
